R Notebook

### 2021-12-01 src/main/r/numeric_vs_hpo_for_liver_function.Rmd
compare raw and hpo of kidney labs for acute kidney failure

## Goal
Compare the performance of raw lab values vs HPO interpretations for acute kidney failure.

The analysis process is very similar to the liver function analysis.

Algorithm:
- Define a cohort of patients coded with acute kidney failure as the primay diagnosis of a visit;
- Get all the lab tests with selected LOINC code; 
- For each patient, the labs within X days (e.g. 7 days) before the diagnosis date is considered to be associated with + outcome; and the labs further outside of the window is considered to be associated with - outcome; 
- For each LOINC code, we will prepare a dataframe, separate into training and test set, train a logistic regression classifier using either raw numerical value or HPO value (observed or not observed), build the ROC curve with the test set, and finally compare the ROC curve to determine whether they are statistically different.

@@

Create a cohort for patients with abnormal liver function diagnoses.
```{r}
sql <- read_file('../sql/acute_kidney_failure_cohort.sql')
print(sql)
n<- dbExecute(dsca, sql)
```

Retrieve liver function diagnoses information for the cohort.
```{r}
sql <- "SELECT * FROM cohort"
df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'acute_kidney_failure_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)

@@

Retrieve liver function lab tests for the cohort. 
```{r}
sql <- read_file('../sql/get_labs_for_acute_kidney_failure.sql')
df_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_acute_kidney_failure.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
df_labs %>% nrow()
```

The first thing is to check how do the raw values of lab test look like before diagnoses of abnormal liver function. We want to choose a time window during which the liver function labs are highly correlated with the diagnosis, and another time window well, which is before the first window and corresponds to a *normal* liver function labs. The plot below shows that roughly we can choose -180 as the time cutoff: within this time cutoff, i.e. (-180, 0) days, the latest lab test tends to be abnormal and correspond to an abnormal liver function diagnosis; outside of this time cutoff, i.e. 180 days before a known liver abnormal function diagnosis, the lab test tends to be normal.  
```{r}
plot_lab_raw_values_vs_rel_date <- function(lab_loinc, ylim=NULL){
  df <- df_labs %>% 
    filter(loinc == lab_loinc) %>% 
    filter(rel_date <= 0 & rel_date > -730) %>% 
    mutate(lab_result_numeric_value = as.numeric(lab_result_numeric_value))

@@

# setClass('LiverLab', representation(lab_loinc = 'character', lab_name = 'character', ylim_min = "numeric", ylim_max = "numeric"))
# liver_lab_ALT <- new('LiverLab', lab_loinc = '1742-6', lab_name = 'Alanine aminotransferase', ylim_min = 0, ylim_max = 500)
# liver_lab_GGT <- new('LiverLab', lab_loinc = '2324-2', lab_name = 'Gamma glutamyl transferase', ylim_min = 0, ylim_max = 1000)
# liver_lab_AST <- new('LiverLab', lab_loinc = '1920-8', lab_name = 'Aspartate aminotransferase', ylim_min = 0, ylim_max = 500)
# liver_lab_ALP <- new('LiverLab', lab_loinc = '6768-6', lab_name = 'Alkaline phosphatase', ylim_min = 0, ylim_max = 1000)
# liver_lab_total_protein <- new('LiverLab', lab_loinc = '2885-2', lab_name = 'Total protein', ylim_min = 0, ylim_max = 15)
# liver_lab_serum_albumin <- new('LiverLab', lab_loinc = '1751-7', lab_name = 'Serum albumin', ylim_min = 0, ylim_max = 20)
# liver_lab_PT <- new('LiverLab', lab_loinc = '5902-2', lab_name = 'Prothrombin time', ylim_min = 0, ylim_max = 50)
# liver_lab_total_bulirubin <- new('LiverLab', lab_loinc = '1975-2', lab_name = 'Total bilirubin', ylim_min = 0, ylim_max = 15)
# liver_lab_direct_bulirubin <- new('LiverLab', lab_loinc = '1968-7', lab_name = 'Direct bilirubin', ylim_min = 0, ylim_max = 15)
# 
# liver_labs <- c(liver_lab_ALT, 
#                 liver_lab_GGT, 
#                 liver_lab_AST, 
#                 liver_lab_ALP, 
#                 liver_lab_total_protein, 
#                 liver_lab_serum_albumin, 
#                 liver_lab_total_bulirubin, 
#                 liver_lab_direct_bulirubin)
# 
# plots <- map(liver_labs, function(x) plot_lab_raw_values_vs_rel_date(lab_loinc = x@lab_loinc, ylim = c(x@ylim_min, x@ylim_max)))
# p <- arrangeGrob(sapply(plots, function(x) x[[1]]), ncol = 2)

p1 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2160-0', ylim = c(0, 20)) + ggtitle('Serum creatinine')

p2 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '3094-0', ylim = c(0, 150)) + ggtitle('Urea nitrogen in serum')

p3 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '6299-2', ylim = c(0, 200)) + ggtitle('Urea nitrogen in blood')

p4 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2823-3', ylim = c(0, 10)) + ggtitle('Potassium in serum')

# seems the test strip only reports positive or negative outcomes
p5 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '5804-0', ylim = c(0, 500)) + ggtitle('protein in urine')

# very sparse data
p6 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2889-4', ylim = c(0, 200)) + ggtitle('Protein in 24h urine')

# very very sparse data
p7 <- plot_lab_raw_values_vs_rel_date(lab_loinc = '2888-6', ylim = c(0, 500)) + ggtitle('Protein in urine')

p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, ncol = 3)

ggsave(filename = 'cache/kidney_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
```

![liver function labs](./cache/kidney_labs_raw_data_distribution_by_rel_date.png)

@@

df_labs_and_outcomes <- df_labs %>% filter(loinc %in% c('2160-0', '3094-0', '6299-2', '2823-3')) %>%
  as_tibble() %>%
  select(-min_diag_date, -rel_date) %>% ## these two columns can be removed from the sql query
  ##filter(medical_record_number %in% c('2715515', '3087585')) %>% ## test with a few patient
  left_join(df_cohort, by = 'medical_record_number') %>%
  mutate(lab_date = ymd(lab_date), min_diag_date = ymd(min_diag_date)) %>%
  mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -180, 0, Inf))) %>%
##select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num)
  filter(window_num %in% c('(-Inf,-180]', '(-180,0]')) %>%
  arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
  group_by(medical_record_number, loinc, window_num) %>%
  mutate(sequence = row_number()) %>%
  #select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num, sequence) %>%
  ungroup() %>%
  filter(sequence == 1) %>%
  mutate(outcome = if_else(window_num == '(-180,0]', 'abnormal', 'normal'))

# the following shows that we have pretty balanced classes for each lab test
df_labs_and_outcomes %>% group_by(loinc, outcome) %>% summarise(n = n())
```

@@

pipeline_compare_for_outcome_prediction <- function(df_one_loinc) {

@@

example_loinc = '2160-0'
df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == example_loinc)
pipeline_compare_for_outcome_prediction(df_one_loinc = df_one_loinc)
```

@@

kidney_analysis <- df_labs_and_outcomes %>%
  group_by(loinc) %>%
  nest() %>%
  mutate(result = future_map(data, pipeline_compare_for_outcome_prediction)) %>%
  select(-data) %>%
  unnest(result)

kidney_analysis
```

```{r}
df_one_loinc <- df_labs_and_outcomes %>% filter(loinc == '1742-6')

>>>>

## Goal
Compare the performance of raw lab values vs HPO interpretations for liver function prediction.

Algorithm:
- Define a cohort of patients coded with abnormal liver function, and the earliest date of diagnosis;
- Get all the lab tests with selected LOINC code; 
- For each patient, the labs within X days (e.g. 7 days) before the diagnosis date is considered to be associated with + outcome; and the labs further outside of the window is considered to be associated with - outcome; 
- For each LOINC code, we will prepare a dataframe, separate into training and test set, train a logistic regression classifier using either raw numerical value or HPO value (observed or not observed), build the ROC curve with the test set, and finally compare the ROC curve to determine whether they are statistically different.

@@

Create a cohort for patients with abnormal liver function diagnoses.
```{r}
sql <- read_file('../sql/liver_abnormal_function_cohort.sql')
print(sql)
n<- dbExecute(dsca, sql)
```

Retrieve liver function diagnoses information for the cohort.
```{r}
sql <- "SELECT * FROM cohort"
df_cohort <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = 'liver_diagnoses.csv', db_access_callback = dbGetQuery, conn = dsca, statement = sql, force_rerun = FALSE)

@@

Retrieve liver function lab tests for the cohort. 
```{r}
sql <- read_file('../sql/get_labs_for_liver_function.sql')
df_liver_labs <- get_df_from_db_or_cache(cache_dir = CACHE_DIR, file_name = "labs_liver_function.csv", db_access_callback = dbGetQuery, conn=dsca, statement = sql, force_rerun = FALSE)
df_liver_labs %>% nrow()
```

The first thing is to check how do the raw values of lab test look like before diagnoses of abnormal liver function. We want to choose a time window during which the liver function labs are highly correlated with the diagnosis, and another time window well, which is before the first window and corresponds to a *normal* liver function labs. The plot below shows that roughly we can choose -180 as the time cutoff: within this time cutoff, i.e. (-180, 0) days, the latest lab test tends to be abnormal and correspond to an abnormal liver function diagnosis; outside of this time cutoff, i.e. 180 days before a known liver abnormal function diagnosis, the lab test tends to be normal.  
```{r}
plot_liver_lab_raw_values_vs_rel_date <- function(lab_loinc, ylim=NULL){
  df <- df_liver_labs %>% 
    filter(loinc == lab_loinc) %>% 
    filter(rel_date <= 0 & rel_date > -730) %>% 
    mutate(lab_result_numeric_value = as.numeric(lab_result_numeric_value))

@@

# setClass('LiverLab', representation(lab_loinc = 'character', lab_name = 'character', ylim_min = "numeric", ylim_max = "numeric"))
# liver_lab_ALT <- new('LiverLab', lab_loinc = '1742-6', lab_name = 'Alanine aminotransferase', ylim_min = 0, ylim_max = 500)
# liver_lab_GGT <- new('LiverLab', lab_loinc = '2324-2', lab_name = 'Gamma glutamyl transferase', ylim_min = 0, ylim_max = 1000)
# liver_lab_AST <- new('LiverLab', lab_loinc = '1920-8', lab_name = 'Aspartate aminotransferase', ylim_min = 0, ylim_max = 500)
# liver_lab_ALP <- new('LiverLab', lab_loinc = '6768-6', lab_name = 'Alkaline phosphatase', ylim_min = 0, ylim_max = 1000)
# liver_lab_total_protein <- new('LiverLab', lab_loinc = '2885-2', lab_name = 'Total protein', ylim_min = 0, ylim_max = 15)
# liver_lab_serum_albumin <- new('LiverLab', lab_loinc = '1751-7', lab_name = 'Serum albumin', ylim_min = 0, ylim_max = 20)
# liver_lab_PT <- new('LiverLab', lab_loinc = '5902-2', lab_name = 'Prothrombin time', ylim_min = 0, ylim_max = 50)
# liver_lab_total_bulirubin <- new('LiverLab', lab_loinc = '1975-2', lab_name = 'Total bilirubin', ylim_min = 0, ylim_max = 15)
# liver_lab_direct_bulirubin <- new('LiverLab', lab_loinc = '1968-7', lab_name = 'Direct bilirubin', ylim_min = 0, ylim_max = 15)
# 
# liver_labs <- c(liver_lab_ALT, 
#                 liver_lab_GGT, 
#                 liver_lab_AST, 
#                 liver_lab_ALP, 
#                 liver_lab_total_protein, 
#                 liver_lab_serum_albumin, 
#                 liver_lab_total_bulirubin, 
#                 liver_lab_direct_bulirubin)
# 
# plots <- map(liver_labs, function(x) plot_liver_lab_raw_values_vs_rel_date(lab_loinc = x@lab_loinc, ylim = c(x@ylim_min, x@ylim_max)))
# p <- arrangeGrob(sapply(plots, function(x) x[[1]]), ncol = 2)

## LOINC 1742-6 Alanine aminotransferase
p1 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1742-6', ylim = c(0, 500)) + ggtitle('Alanine aminotransferase')

## LOINC 2324-2 Gamma glutamyl transferase
p2 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '2324-2', ylim = c(0, 1000)) + ggtitle('Gamma glutamyl transferase')

## LOINC 1920-8 Aspartate aminotransferase
p3 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1920-8', ylim = c(0, 500)) + ggtitle('Aspartate aminotransferase')

## LOINC 6768-6 Alkaline phosphatase
p4 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '6768-6', ylim = c(0, 1000)) + ggtitle('Alkaline phosphatase')

## LOINC 2885-2 total protein: if low, maybe reflecting liver or kidney problem
p5 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '2885-2', ylim = c(0, 15)) + ggtitle('Total protein')

## LOINC 1751-7 serum albumin
p6 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1751-7', ylim = c(0, 7.5)) + ggtitle('Serum albumin')

## LOINC 5902-2 Prothrombin time
p7 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '5902-2', ylim = c(0, 50)) + ggtitle('PT')

## LOINC 1975-2 Bilirubin total
p8 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1975-2', ylim = c(0, 15)) + ggtitle('Bilirubin total')

## LOINC 1968-7 Bilirubin direct
p9 <- plot_liver_lab_raw_values_vs_rel_date(lab_loinc = '1968-7', ylim = c(0, 15)) + ggtitle('Bilirubin direct')

p <- gridExtra::arrangeGrob(p1, p2, p3, p4, p5, p6, p7, p8, p9, ncol = 3)

ggsave(filename = 'cache/liver_labs_raw_data_distribution_by_rel_date.png', plot = p, width = 6, height = 6)
```

![liver function labs](./cache/liver_labs_raw_data_distribution_by_rel_date.png)

Machine learning part:
  For each LOINC lab test
    prepare dataframe for each 
    data split
    train logistic regression with either raw numeric value or HPO terms
    build ROC with test set
    run statistical comparison to compare the two models

@@

df_liver_labs_and_outcomes <- df_liver_labs %>% 
  as_tibble() %>%
  select(-min_diag_date, -rel_date) %>% ## these two columns can be removed from the sql query
  ##filter(medical_record_number %in% c('2715515', '3087585')) %>% ## test with a few patient
  left_join(df_cohort, by = 'medical_record_number') %>%
  mutate(lab_date = ymd(lab_date), min_diag_date = ymd(min_diag_date)) %>%
  mutate(lab_date_rel_to_diag = as.duration(interval(min_diag_date, lab_date)) / ddays()) %>%
  mutate(window_num = cut(lab_date_rel_to_diag, c(-Inf, -180, 0, Inf))) %>%
##select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num)
  filter(window_num %in% c('(-Inf,-180]', '(-180,0]')) %>%
  arrange(medical_record_number, loinc, -lab_date_rel_to_diag) %>%
  group_by(medical_record_number, loinc, window_num) %>%
  mutate(sequence = row_number()) %>%
  #select(medical_record_number, loinc, lab_date, min_diag_date, lab_date_rel_to_diag, window_num, sequence) %>%
  ungroup() %>%
  filter(sequence == 1) %>%
  mutate(outcome = if_else(window_num == '(-180,0]', 'abnormal', 'normal'))

# the following shows that we have pretty balanced classes for each lab test
df_liver_labs_and_outcomes %>% group_by(loinc, outcome) %>% summarise(n = n())
```

@@

pipeline_compare_for_liver_function <- function(df_one_loinc) {

@@

example_loinc = '1742-6'
df_one_loinc <- df_liver_labs_and_outcomes %>% filter(loinc == example_loinc)
pipeline_compare_for_liver_function(df_one_loinc = df_one_loinc)
```

@@

liver_analysis <- df_liver_labs_and_outcomes %>%
  group_by(loinc) %>%
  nest() %>%
  mutate(result = future_map(data, pipeline_compare_for_liver_function)) %>%
  select(-data) %>%
  unnest(result)

liver_analysis
```

```{r}
df_one_loinc <- df_liver_labs_and_outcomes %>% filter(loinc == '1742-6')
